Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-29 submission/manuscript.Rmd
discussion CGI drafts

@@

- CpGme was imparied at some germline genes, including ones expressed throughout life
-  Some regions, such as Dazl, were able to compensate for KDM5C loss, even though it gained H3K4me2 in EpiLCs. 
	- Multiple repressive mechanisms
We found _Kdm5c_-KO EpiLCs and the amygdala accumulate H3K4me2/3 around the TSS of germline genes, potential indicating KDM5C's cataltyic activity is necessary at later points in development.

KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as

as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters. Our study therefore indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.

@@

- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms

>>>>

- Our results indicate KDM5C's removal of H3K4me2/3 may be necessary at a subset of CGI-containing germline genes, given they fail to accumulate CpGme during EpiLC differentiation. These CpGme-sensitive germline genes, such as _D1Pas1_ and _Naa11_, are also de-repressed in the mature _Kdm5c_-KO brain and consistently dysregulated in multiple _Kdm5c_-KO datasets[@scandagliaLossKdm5cCauses2017].

@@

- Contrastingly, many germline genes had only a modest reduction in CpGme, including _Dazl_ even though its TSS accumulated substantial H3K4me2. 
- Multiple regulators repress dazl, dazl is not expressed in the mature _Kdm5c_-KO brain
	E2F6/MGA/MAX
		- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
		-  Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci

- Altogether, our results indicates a given chromatin modifier's catalytic and non-catlytic regulation of gene expression can change depending upon the locus and developmental stage.

Although the classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter. We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS.

- CGIs highly determined KDM5C recruitment to germline gene promoters.

- KDM5C loss also impacted CpGme at some KDM5C-unbound, CGI-free germline genes. Unclear what the repressive mechanism is for these genes.

In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.

- Germline gene suppression has focused on the accumulation of CpGme at CGIs

- DNAme and CpG islands

@@

- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms
